---
title: Characteristic Function
author: Erik-Jan van Kesteren
description: "Visualising the characteristic function and eigenvalues of a matrix"
output: 
  html_document:
    theme: spacelab
---

```{r setup, include=FALSE, echo=FALSE,render=FALSE,warning=FALSE,background=TRUE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(dev = 'svg')
```

## Packages needed for this blog
```{r pkg, echo=TRUE,warning=FALSE,background=TRUE, message=FALSE}
library(ggplot2)
library(manipulate)
library(firatheme) # https://github.com/vankesteren/firatheme
library(Massign)   # https://github.com/vankesteren/Massign
```

## The characteristic function

A while ago, I blogged that a matrix $A$ can be seen as an [operation](../transformation-matrix) and that the determinant of this matrix $|A|$ says something about the volume of the transformation. This blog is about another property of matrices: the characteristic function.

The characteristic function of a square matrix $A$ of order $n$ is defined as follows:

$$p_A(\lambda) = | \lambda I_n - A |$$

Let's look at this function for the following $2\times 2$ matrix:

$$A = \begin{bmatrix} 3 & 1 \\ 2 & 4  \end{bmatrix}$$
$$ \begin{align} 
p_A(\lambda) &= | \lambda I_n - A | \\
&= \left| \begin{bmatrix} \lambda & 0 \\ 0 & \lambda \end{bmatrix} - \begin{bmatrix} 3 & 1 \\ 2 & 4  \end{bmatrix} \right| \\
&= \left| \begin{bmatrix} \lambda - 3 & -1 \\ -2 & \lambda - 4  \end{bmatrix} \right|\\
\end{align}$$

The roots of this function - a polynomial of order $n$ - are the eigenvalues of the matrix. We can find them using some algebra, remembering that the determinant of a $2\times 2$ matrix is calculated as $ad-bc$:

$$ \begin{align}
p_A(\lambda) &= (\lambda - 3)(\lambda - 4) + 2\\
&= \lambda^2 - 7\lambda + 10 \\
&= (\lambda - 2)(\lambda - 5)
\end{align}$$

So $\lambda = 2$ or $\lambda = 5$. These are the two eigenvalues of this matrix.

## Visualising the characteristic function

Using the power of `R`, we can get a better intuition for the characteristic function by visualising it. Below the code for a function that takes a matrix and visualises this function.

```{r, echo=TRUE}
charfun <- function(mat, from, to) {
  n <- ncol(mat)
  stopifnot(n == nrow(mat))
  x <- seq(from, to, length.out = 3000)
  
  ggdat <- data.frame(
    x = x,
    y = vapply(x, function(lambda) det(lambda*diag(n) - mat), 1.0)
  )
  ev <- eigen(mat)$values
  ev <- ev[ev <= to & ev > from]
  
  evdat <- data.frame(x = ev, y = rep(0, length(ev)))
  
  ggplot(ggdat, aes(x = x, y = y)) + 
    geom_hline(yintercept = 0, col = firaCols[5], lwd = 1) +
    geom_vline(xintercept = ev, col = firaCols[2], lwd = 1, lty = 2) +
    geom_line(col = firaCols[1], lwd = 1) + 
    geom_point(aes(x, y), evdat, size = 3, col = firaCols[1]) +
    labs(x = "Lambda", 
         y = "Characteristic function value", 
         title = "Characteristic function of a matrix") +
    theme_fira()
}

A %<-% "3, 1
        2, 4"

charfun(A, 0, 7) + ggtitle("Characteristic function of A")
```

## What can we see?
1. The characteristic function crosses the axis at 2 and 5, just as we expected
2. The characteristic function is indeed a quadratic function, i.e., a polynomial of order $n = 2$.

You can play around with this function in `R` by trying out different matrices. Try out a covariance matrix, different kinds of symmetric and assymmetric matrices!

## Bonus: interactive
After running the above `R` chunks, you can run the following to play around with different covariance matrices of the following form:

$$A = \begin{bmatrix} 1 & a & b \\ a & 1 & c \\ b & c & 1 \end{bmatrix}$$
Play around with it to see what happens to the eigenvalues of this matrix! For example, note that when a, b, c, and d are all 0 the eigenvalues are all 1. There are some nice symmetries to be explored here.

```{r, echo=TRUE, warning=FALSE, eval=FALSE}
manipulate(
  {
    A %<-% " 1,
             a,  1
             b,  c,  1"
    charfun(A, 0, 2) + ylim(c(-.7,.7))
  },
  a = slider(-1, 1, initial = .5, step = 0.1),
  b = slider(-1, 1, initial = .3, step = 0.1),
  c = slider(-1, 1, initial = .2, step = 0.1)
)
```


```{r, echo=FALSE}
A %<-% " 1,
        .5,  1
        .3,  .2,  1"
charfun(A, 0, 2) + ylim(c(-.7,.7))
```