---
title: "Why would you do regression with multivariate outcome $Y$?"
author: "Erik-Jan van Kesteren"
format: 
  html:
    toc: true
    self-contained: true
    self-contained-math: true
    html-math-method: katex
---

A question that's been bothering me for a while is the following: if you have $Q$ outcome variables in a matrix, say $Y \in \mathbb{R}^{N\times Q}$, and you want to regress them on a common set of $P$ predictors, say $X \in \mathbb{R}^{N\times P}$, what is the advantage of doing this in one model $Y = XB + E$ as opposed to $Q$ separate models? My intuition is that if the outcome variables in $Y$ (or rather the errors $E$) are correlated, this will lead to smaller standard errors for the parameters.

Let's see if we can confirm or deny this intuition with a small simulation.

## Generating data

First, we create a function to generate some data following the above pattern. There are a few specific parameters in this function, the most important are `varprop` (the proportion of variance in Y explained by X) and the various `rho` parameters (which tune the correlations of `B`, `X` and `Y`). 

::: {.callout-note}
Code blocks which are not the most relevant for following the content are collapsed. Click the "code" button to show the code for those sections.
:::

```{r}
#| label: datasim
#| code-fold: true

#' Function for simulating multivariate regression data
#' 
#' @param N the number of observations
#' @param P the number of predictors
#' @param Q the number of outcome variables
#' @param varprop the proportion of variance explained in each outcome
#' @param rho_X the correlation between the predictors
#' @param rho_B the correlation between the parameters
#' @param rho_E the correlation between the errors
#' @param exact whether the error covariance should be imposed exactly (TRUE)
#' 
#' @return list with Y, X
sim_dat <- function(N = 100, P = 5, Q = 2, varprop = 0.5, rho_X = 0, rho_B = 0, rho_E = 0, exact = TRUE) {
  # Create X matrix
  X <- matrix(rnorm(N*P), N)
  RX <- matrix(rho_X, P, P)
  diag(RX) <- 1
  CRX <- chol(RX)
  X <- if (exact) X %*% solve(chol(var(X)), CRX) else X %*% CRX
  
  # Create B matrix
  B <- matrix(rnorm(P*Q), P)
  RB <- matrix(rho_B, Q, Q)
  diag(RB) <- 1
  CRB <- chol(RB)
  B <- if (exact) B %*% solve(chol(var(B)), CRB) else B %*% CRB
  
  # Create E matrix
  E <- matrix(rnorm(N*Q), N)
  XB <- X %*% B
  vXB <- diag(var(XB))
  sE <- sqrt(vXB*varprop/(1-varprop))
  RE <- matrix(rho_E, Q, Q)
  diag(RE) <- 1
  CVE <- chol(RE) %*% diag(sE)
  E <- if (exact) E %*% solve(chol(var(E)), CVE) else E %*% CVE
  
  # Create Y matrix
  Y <- XB + E
  
  return(list(Y = Y, X = X, B = B, E = E))
}
``` 

The function returns a list with all the generated matrices and can be used as follows:

```{r}
#| label: sim
#| eval: false
dat <- sim_dat(
  N = 100, P = 5, Q = 2, varprop = 0.5,
  rho_X = 0, rho_B = 0, rho_E = 0, exact = TRUE
)

dat$X
dat$Y
```

## Separate vs. joint model

First, let's estimate the separate regressions and the multivariate one when the errors are uncorrelated and compare the standard errors
```{r}
#| label: compare_nocor
# set seed and generate data
set.seed(45)
d_nocor <- sim_dat()

# fit separate regressions
fit_sep_nocor_1 <- lm(Y[,1] ~ X + 0, data = d_nocor)
fit_sep_nocor_2 <- lm(Y[,2] ~ X + 0, data = d_nocor)

# fit multivariate regression
fit_joint_nocor <- lm(Y ~ X + 0, data = d_nocor)
```
::: {.panel-tabset}
## Estimates of separate regressions
```{r}
#| label: nocor1
#| code-fold: true
#| results: hold
# extract coef & s.e.
cat("Separate models:\n")
rbind(
  summary(fit_sep_nocor_1)$coefficients[,1:2], 
  summary(fit_sep_nocor_2)$coefficients[,1:2]
)
```

## Estimates of joint regression
```{r}
#| label: nocor2
#| code-fold: true
#| results: hold
# extract coef & s.e.
cat("Joint model:\n")
rbind(
  summary(fit_joint_nocor)[[1]]$coefficients[,1:2],
  summary(fit_joint_nocor)[[2]]$coefficients[,1:2]
)
```
:::
In this case, it does not matter at all whether you perform two separate regressions or whether you perform the regressions together. 

## Correlated errors

Now, let's do the same thing but with residual correlations of 0.4:

```{r}
#| label: compare_corE

# set seed and generate data
set.seed(45)
d_cor_E <- sim_dat(rho_E = 0.4)

# fit separate regressions
fit_sep_cor_1 <- lm(Y[,1] ~ X + 0, data = d_cor_E)
fit_sep_cor_2 <- lm(Y[,2] ~ X + 0, data = d_cor_E)

# fit multivariate regression
fit_joint_cor <- lm(Y ~ X + 0, data = d_cor_E)
```

::: {.panel-tabset}
## Estimates of separate regressions
```{r}
#| label: cor1
#| code-fold: true
#| results: hold
# extract coef & s.e.
cat("Separate models:\n")
rbind(
  summary(fit_sep_cor_1)$coefficients[,1:2], 
  summary(fit_sep_cor_2)$coefficients[,1:2]
)
```

## Estimates of joint regression
```{r}
#| label: cor2
#| code-fold: true
#| results: hold
# extract coef & s.e.
cat("Joint model:\n")
rbind(
  summary(fit_joint_cor)[[1]]$coefficients[,1:2],
  summary(fit_joint_cor)[[2]]$coefficients[,1:2]
)
```
:::

Even in this case, there is no difference between the two!


::: {.callout-note}
This is really counterintuitive to me! Can't we "borrow information" somehow in the joint model? Surely there are differences in the asymptotic covariance matrix of $B$?
:::

Let's inspect asymptotic covariance matrix (ACOV) of the separate regressions, which is block-diagonal, whereas the ACOV of the joint model is not:

```{r}
#| label: computeacov
#| code-fold: true
#| warning: false
#| message: false
library(Matrix)

ACOV_joint <- vcov(fit_joint_cor)
ACOV_sep <- as.matrix(bdiag(vcov(fit_sep_cor_1), vcov(fit_sep_cor_2)))
rownames(ACOV_sep) <- 
  colnames(ACOV_sep) <- 
  rownames(ACOV_joint) <- 
  colnames(ACOV_joint) <- 
  c(outer(paste0("Y", 1:2), paste0("X", 1:5), paste, sep = "<-"))
```

::: {.panel-tabset}
## ACOV of separate regressions
```{r}
#| label: acov1
#| echo: false

round(ACOV_sep, 3)
```
## ACOV of joint regression
```{r}
#| label: acov2
#| echo: false
round(ACOV_joint, 3)
```
:::

So indeed, there is a difference in the asymptotic covariance matrices of the parameters in the separate and joint case, but this does not make a difference for the standard errors.

## Another option?

Maybe inducing correlations in some the other data matrices makes a difference? Let's check:
```{r}
#| label: compare_corB
# set seed and generate data
set.seed(45)
d_cor_all <- sim_dat(rho_B = 0.4, rho_E = 0.4, rho_X = 0.4)

# fit separate regressions
fit_sep_cor_1 <- lm(Y[,1] ~ X + 0, data = d_cor_all)
fit_sep_cor_2 <- lm(Y[,2] ~ X + 0, data = d_cor_all)

# fit multivariate regression
fit_joint_cor <- lm(Y ~ X + 0, data = d_cor_all)

```


::: {.panel-tabset}
## Estimates of separate regressions
```{r}
#| label: corb1
#| code-fold: true
#| results: hold
# extract coef & s.e.
cat("Separate models:\n")
rbind(
  summary(fit_sep_cor_1)$coefficients[,1:2], 
  summary(fit_sep_cor_2)$coefficients[,1:2]
)

```

## Estimates of joint regression
```{r}
#| label: corb2
#| code-fold: true
#| results: hold
# extract coef & s.e.
cat("Joint model:\n")
rbind(
  summary(fit_joint_cor)[[1]]$coefficients[,1:2],
  summary(fit_joint_cor)[[2]]$coefficients[,1:2]
)
```
:::

Also no difference!


## Seemingly unrelated regressions

There is another way of analysing this data, and it goes like so: stack the $Y$ variables on top of each other, and create a block diagonal matrix for $X$. Then, allow for cross-correlations between the residuals. See [the SUR wiki page](https://en.wikipedia.org/wiki/Seemingly_unrelated_regressions). 

::: {.callout-note}
SUR is actually meant to deal with varying predictors for the different equations, so this is not exactly the intended use-case! In the case of different predictors for each of the equations, there would be a difference between SUR and OLS ([see here](https://en.wikipedia.org/wiki/Seemingly_unrelated_regressions#Equivalence_to_OLS))
:::

```{r}
#| label: SUR
#| message: false
#| warning: false

library(systemfit)
eq_list <- list(
  Y1 = Y[,1] ~ X + 0,
  Y2 = Y[,2] ~ X + 0
)
summary(systemfit(eq_list, method = "SUR", data = d_cor_all))
```

As we can see, this again gives exactly the same parameter estimates and standard errors.

## Conclusion

So why would you do regression with multivariate outcome $Y$? If you have the same set of predictors for each outcome, you wouldn't. Just do separate regressions and call it a day! Also: intuitions are not always correct 🙂


