---
title: "Efficient Kullback-Leibler divergence of two Gaussian distributions in R"
author: "Erik-Jan van Kesteren"
format: 
  html:
    toc: true
    self-contained: true
    self-contained-math: true
    html-math-method: katex
---

## Introduction
In this document, I create a small `R` function for computing the Kullback-Leibler divergence between two (multivariate) normal (Gaussian) distributions. The general formula for the divergence is as follows:

$$D_{KL}(P \parallel Q) = \int_{-\infty}^{\infty} p(x) \log \left(\frac{p(x)}{q(x)}\right)\, dx$$
If we assume that both $p(x)$ and $q(x)$ are multivariate normal, then all the difficult integration drops out and we are left with the following
formula, adapted from [this excellent answer on StackOverflow](https://stats.stackexchange.com/a/60699/116878):

$$D_{KL}(P \parallel Q) = \frac{1}{2}\left[\log|\Sigma_q|-\log|\Sigma_p| - d + \text{tr} \{ \Sigma_q^{-1}\Sigma_p \} + (\mu_q - \mu_p)^T \Sigma_q^{-1}(\mu_q - \mu_p)\right] $$
Where $d$ indicates the dimensionality of the distributions, and $|\cdot|$ indicates the determinant.

Of course, a function already exists for computing this divergence in `R`. Specifically, it exists as [`rags2ridges::KLdiv()`](https://search.r-project.org/CRAN/refmans/rags2ridges/html/KLdiv.html). However, when installing the `{rags2ridges}` package, several dependencies need to be pulled from bioconductor as they are not on CRAN, and the package in general has a lot of functionality we might not need if we just want to compute the KL-divergence.

The goal of this post is thus to create a light-weight, performant version of the KL-divergence function which is fast, has low memory requirements, and no dependencies beyond what's included in base `R`. Below, I show how I achieved a big speedup compared to the existing `KLdiv` function.

## Example inputs

We will be using the following example inputs:
```{r}
#| label: example inputs
set.seed(45)

# we use 25-dimensional normal distributions
P <- 25

# mean vectors
mean_0 <- rnorm(P)
mean_1 <- rnorm(P)

# covariance matrices
S_0 <- rWishart(1, P*2, diag(P))[,,1]
S_1 <- rWishart(1, P*2, diag(P))[,,1]
```


The existing function returns the following value for $D_{KL}(P \parallel Q)$ with these inputs (note that the ordering is a bit weird; KL-divergence is not symmetric so order is important):

```{r}
#| label: existing kldiv
rags2ridges::KLdiv(mean_1, mean_0, S_1, S_0)
```

## Computing log-determinants
The first two terms in our equation are the log-determinants of the covariance matrices:

$$\log|\Sigma_q|-\log|\Sigma_p|$$

There are several ways to compute a log-determinant in `R`; here, we benchmark three methods, once on the 25-dimensional `S_1` we created earlier, and another time on a random 1000-dimensional covariance matrix:

```{r}
#| label: log-determinants

bench::mark(
  logdet = log(det(S_1)),
  determ = determinant(S_1)$modulus,
  cholesky = 2*sum(log(diag(chol(S_1)))),
  check = FALSE
)

set.seed(45)
S1000 <- rWishart(1, 1000, diag(1000))[,,1]
bench::mark(
  logdet = log(det(S1000)),
  determ = determinant(S1000)$modulus,
  cholesky = 2*sum(log(diag(chol(S1000)))),
  check = FALSE
)
```

Generally, `determinant()` performs slightly better than `log(det())`. For low-dimensional distributions, the `cholesky` version is worst, but it is actually fastest for the 1000-dimensional distribution. This is especially interesting as  we may be able to reuse this cholesky decomposition to speed up later steps.

## Computing the trace
The second difficult term to compute is the trace part:

$$\text{tr} \{ \Sigma_q^{-1}\Sigma_p \}$$

Note that here we need the inverse of $\Sigma_q$, which we will need at a later stage as well. Here, we again have several options for computing the trace. One trick is adapted from the [`lavaan`](https://lavaan.org) package, from the code [here](https://github.com/yrosseel/lavaan/blob/4eb699f1300d22657041f1d04965a8cb6d89811f/R/lav_matrix.R#L970-L974), with the additional knowledge that covariance matrices are symmetric.

```{r}
#| label: trace

bench::mark(
  naive = sum(diag(solve(S_1) %*% S_0)),
  solve = sum(diag(solve(S_1, S_0))),
  lavaan = sum(solve(S_1) * S_0)
)
```

Here, the naive method is definitely the worst, and the other two methods are almost at the same level. Again, the last identity is especially interesting, because we want to precompute $\Sigma_q^{-1}$ anyway for the next part.

Note that if we assume that we have the cholesky decomposition of $\Sigma_q$ precomputed from the previous part, this equation -- especially the last one -- becomes even faster:

```{r}
#| label: trace chol

S_1_c <- chol(S_1)

bench::mark(
  naive = sum(diag(chol2inv(S_1_c) %*% S_0)),
  solve = sum(diag(forwardsolve(S_1_c, backsolve(S_1_c, S_0, transpose = TRUE), upper.tri = TRUE))),
  lavaan = sum(chol2inv(S_1_c) * S_0)
)
```

## Quadratic form
The last computationally intensive part of the equation is the following quadratic form: 

$$(\mu_q - \mu_p)^T \Sigma_q^{-1}(\mu_q - \mu_p)$$

Here, we first compute the mean differences, which we call `delta`. Then, we again need the inverse of the covariance matrix of $q(x)$, so we will assume that it is pre-computed from the previous step.

```{r}
#| label: quadratic form

delta <- mean_1 - mean_0
Omega_1 <- chol2inv(S_1_c)

bench::mark(
  naive = t(delta) %*% Omega_1 %*% delta,
  cross = crossprod(delta, Omega_1 %*% delta),
  cross2 = crossprod(delta, crossprod(Omega_1, delta))
)
```

The second version, using only one `crossprod()` is the fastest here. It's actually quite significantly faster than the naïve implementation with standard matrix products.

## Putting it all together {#sec-optimized}

Combining all the above sections, we can make two versions of the KL-divergence function. One which precomputes and uses the Cholesky decomposition for $\Sigma_q$, and one which does not. We can compare them for our example distributions:

```{r}
#| label: kldiv

kldiv_base <- function(mu_p, mu_q, Sigma_p, Sigma_q) {
  logdet_p <- determinant(Sigma_p)$modulus
  logdet_q <- determinant(Sigma_q)$modulus
  d <- length(mu_p)
  Omega_q <- solve(Sigma_q)
  trace <- sum(Omega_q * Sigma_p)
  delta <- mu_q - mu_p
  quad <- crossprod(delta, Omega_q %*% delta)
  return(c(logdet_q - logdet_p - d + trace + quad) / 2)
}

kldiv_chol <- function(mu_p, mu_q, Sigma_p, Sigma_q) {
  chol_q <- chol(Sigma_q)
  logdet_p <- determinant(Sigma_p)$modulus
  logdet_q <- 2*sum(log(diag(chol_q)))
  d <- length(mu_p)
  Omega_q <- chol2inv(chol_q)
  trace <- sum(Omega_q * Sigma_p)
  delta <- mu_q - mu_p
  quad <- crossprod(delta, Omega_q %*% delta)
  return(c(logdet_q - logdet_p - d + trace + quad) / 2)
}

bench::mark(
  base = kldiv_base(mean_0, mean_1, S_0, S_1),
  chol = kldiv_chol(mean_0, mean_1, S_0, S_1)
)
```

As you can see, the cholesky version is quite a bit (>1/3) faster. But does it also give the correct answer?

```{r}
#| label: answer

kldiv_chol(mean_0, mean_1, S_0, S_1)

```
It does! So let's use that version to compare to the original, existing `rags2ridges::KLdiv()` function. 

```{r}
#| label: comparison

bench::mark(
  ours = kldiv_chol(mean_0, mean_1, S_0, S_1),
  theirs = rags2ridges::KLdiv(mean_1, mean_0, S_1, S_0),
  relative = TRUE
)
```

We've achieved quite a dramatic speedup, between 14 and 15x. Additionally, we have allocated 5x less memory. This speedup is maintained for 2-dimensional distributions:

```{r}
#| label: 2d comparison

# generate parameters
set.seed(45)
P <- 2
mean_0 <- rnorm(P)
mean_1 <- rnorm(P)
S_0 <- rWishart(1, P*2, diag(P))[,,1]
S_1 <- rWishart(1, P*2, diag(P))[,,1]

bench::mark(
  ours = kldiv_chol(mean_0, mean_1, S_0, S_1),
  theirs = rags2ridges::KLdiv(mean_1, mean_0, S_1, S_0),
  relative = TRUE
)
```

And a positive side-effect of the new function and its memory efficiency is that it also works for larger 250-dimensional distributions, unlike the old function:

```{r}
#| label: 250d comparison

# generate parameters
set.seed(45)
P <- 250
mean_0 <- rnorm(P)
mean_1 <- rnorm(P)
S_0 <- rWishart(1, P*2, diag(P))[,,1]
S_1 <- rWishart(1, P*2, diag(P))[,,1]

kldiv_chol(mean_0, mean_1, S_0, S_1)
rags2ridges::KLdiv(mean_1, mean_0, S_1, S_0)

```

## Clean-up and document
What remains is to tidy up the function, make it more robust, and to document it properly. Allow me to present the fast and furious Gaussian Kullback-Leibler divergence:

```{r}
#| label: KL-divergence

#' Kullback-Leibler divergence between two Gaussians
#' 
#' This function computes $D_{KL}(p \parallel q)$, where $p(x)$ 
#' and $q(x)$ are two multivariate normal distributions.
#' 
#' @param mu_p the mean vector of p
#' @param mu_q the mean vector of q
#' @param Sigma_p the covariance matrix of p
#' @param Sigma_q the covariance matrix of q
#' 
#' @return Kullback-leibler divergence from p to q 
#' (numeric scalar)
kldiv <- function(mu_p, mu_q, Sigma_p, Sigma_q) {
  chol_q   <- chol(Sigma_q)
  logdet_p <- determinant(matrix(Sigma_p))$modulus
  logdet_q <- 2*sum(log(diag(chol_q)))
  d        <- length(mu_p)
  Omega_q  <- chol2inv(chol_q)
  trace    <- sum(Omega_q * Sigma_p)
  delta    <- mu_q - mu_p
  quad     <- crossprod(delta, Omega_q %*% delta)
  return(c(logdet_q - logdet_p - d + trace + quad) / 2)
}
```


## Symmetric version

There is also a symmetric version of the Kullback-Leibler divergence, also known as Jeffreys divergence:

$$D_{J}(P, Q) = D_{KL}(P \parallel Q) + D_{KL}(Q \parallel P)$$

We can build on the speedups from before by also precomputing the Cholesky decomposition for $\Sigma_p$. There are some additional properties of the symmetric divergence which are nice for computation:

- the log-determinant terms cancel out, so they do not need to be computed at all
- we can take $d$ out of the inner sum because it enters the equation twice (then we don't need to multiply and divide by 2)
- Because $\text{Tr}\{A\}+\text{Tr}\{B\} = \text{Tr}\{A+B\}$ we can compute the trace at once
- Because `delta` is used only in quadratic forms, we don't need to compute it twice ($\delta^T\delta = (-\delta)^T(-\delta)$
- Then, we can combine the quadratic forms as follows: $x^TAx + x^TBx = x(A+B)x^T$

```{r}
#| label: J-divergence

#' Jeffreys divergence between two Gaussians
#' 
#' This function computes Jeffres divergence, the symmetric 
#' version of the Kullback-Leibler divergence: 
#' $D_{J}(p, q) = D_{KL}(P \parallel Q) + D_{KL}(Q \parallel P)$, 
#' where $p(x)$ and $q(x)$ are two multivariate normal distributions.
#' 
#' @param mu_p the mean vector of p
#' @param mu_q the mean vector of q
#' @param Sigma_p the covariance matrix of p
#' @param Sigma_q the covariance matrix of q
#' 
#' @return Kullback-leibler divergence from p to q 
#' (numeric scalar)
jefdiv <- function(mu_p, mu_q, Sigma_p, Sigma_q) {
  chol_p   <- chol(Sigma_p)
  chol_q   <- chol(Sigma_q)
  d        <- length(mu_p)
  Omega_p  <- chol2inv(chol_p)
  Omega_q  <- chol2inv(chol_q)
  trace    <- sum(Omega_p * Sigma_q + Omega_q * Sigma_p)
  delta    <- mu_q - mu_p
  quad     <- delta %*% (Omega_p + Omega_q) %*% delta
  return(c(trace + quad) / 2 - d)
}
```

When comparing the performance to that of the `{rags2ridges}` package, we see that the speed-up is slightly bigger than with the assymmetric KL-divergence, around 17x.
```{r}
#| label: r2r
set.seed(45)
P <- 25
mean_0 <- rnorm(P)
mean_1 <- rnorm(P)
S_0 <- rWishart(1, P*2, diag(P))[,,1]
S_1 <- rWishart(1, P*2, diag(P))[,,1]

bench::mark(
  ours = jefdiv(mean_0, mean_1, S_0, S_1),
  theirs = rags2ridges::KLdiv(mean_1, mean_0, S_1, S_0, TRUE), 
  relative = TRUE
)
```


## Julia version

We can compare this to the notoriously fast `Julia` programming language. The implementation there needs less customization and can proceed more or less exactly equal to the mathematical formula. Thus, we "simply" write out the function as follows:

```{julia}
#| label: julia version

using MKL
using LinearAlgebra
using BenchmarkTools

# kldiv function
function kl_div(μ_p::Vector, μ_q::Vector, Σ_p::AbstractMatrix, Σ_q::AbstractMatrix)
    d   = length(μ_p)
    Ω_q = inv(Σ_q)
    tr  = sum(Ω_q .* Σ_p)
    δ   = μ_q .- μ_p
    qfm = dot(δ, Ω_q, δ)

    return (logdet(Σ_q) - logdet(Σ_p) - d + tr + qfm) / 2
end
```
Let's check that it has the same output:
```{julia}
#| label: kl-test

# get example inputs from R
mean_0 = @rget mean_0;
mean_1 = @rget mean_1;
S_0 = @rget S_0;
S_1 = @rget S_1;

kl_div(mean_0, mean_1, S_0, S_1)
```
Now, let's benchmark this function:
```{julia}
#| label: benchmark
@benchmark kl_div($mean_0, $mean_1, $S_0, $S_1)
```

The `Julia` function is even faster than the optimized `R` function, completing in less than 20 microseconds (median), whereas the median of the `R` Cholesky version (@sec-optimized) is around 40 microseconds.
